### 2016-10-04 Paper_v7.tex
DB edits 4

The predictive power of the ensemble was tested using both cross validation and by comparing simulations with data sets not used for model training.
In whole, 72\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
generated from a random starting point. Conversely, we \emph{predicted} approximately 72\% of the training objectives, at a 95\% confidence interval compared to randomized parameters.
The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 (not used for model training) to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L). Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
A listing of objective function values resulting from training, cross validation and the random parameter control is given in the supplement (Fig. \ref{fg:ObjTable}).

@@

\subsection*{TGF$-\beta$1/2 and VEGF-A exposure promotes phenotype heterogeneity through NFATc and phosphorylated Sp1}
While we captured the central tendency of many of the molecular features of EMT induction following TGF$-\beta$1/2 exposure, an often neglected but important emergent
feature of developmental and pathological programs is population heterogeneity \citep{Park:2010fu}.
We (and others) have previously hypothesized that deterministic model ensembles can simulate population behavior, at least at a course grained level \citep{Lequieu:2011fj}.
We tested this hypothesis by analyzing the response of the population of EMT models to extracellular cues and then comparing this response to flow cytometry studies.
We quantified the phenotypic response of the individual members of the ensemble to TGF$-\beta$1/2 stimulation for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial) following the addition of TGF$-\beta$1/2 alone (Fig. \ref{fg:F3}), and VEGF-A in combination with NFATc inhibitors (Fig. \ref{fg:F3}).
%We were further able to use this analysis to examine the activity and effects of underlying signaling molecules.

We identified model subpopulations that exhibited different behaviors following exposure to TGF$-\beta$1/2 (Fig. \ref{fg:F3}A).
Analysis of the molecular signatures of these subpopulations suggested the abundance, localization and state of the Sp1, AP-1 and NFATc transcription factors controlled population heterogeneity.
The majority of models ($>$80\%) responded to treatment, moving off of the untreated population (Fig \ref{fg:F3}A-C, gray).
These models showed the classically expected behavior, a switch from an epithelial to mesenchymal phenotype following TGF$-\beta$1/2 exposure.
Models near resembling untreated cells had evaluated nuclear localized phosphorylated Sp1, relative to non-induced cells.
Elevated Sp1 activity decreased E-cadherin expression through Slug-mediated inhibition, which in turn increased Vimentin expression through TGF$-\beta$3 autocrine signaling and the liberation of $\beta$-catenin. However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (black arrow), but also present in a small percentage of untreated cells (Fig. \ref{fg:F3}A).
Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in the simultaneous increase of \emph{both} Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).
Analysis of these hypothetical cells suggested they had deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression.

To test this hypothesis, we simulated the response of the network to TGF$-\beta$1/2 and VEGF-A treatment with and without NFATc inhibitors (Fig. \ref{fg:F4}).
As expected, stimulation with VEGF-A (50 a.u.) maintained an epithelial population, while TGF$-\beta$1/2 (10 a.u.) exposure shifted
the population from an epithelial to a mesenchymal phenotype (Fig. \ref{fg:F3}A and Fig. \ref{fg:F3}B).
On the other hand, combined stimulation with TGF$-\beta$1/2 (10 a.u.) and VEGF-A (50 a.u.) increased both E-cadherin and Vimentin expression, resulting in a hybrid phenotype with both epithelial and mesenchymal characteristics (Fig. \ref{fg:F3}C).
Vimentin expression was correlated with high levels of nuclear phosphorylated Sp1, following TGF$-\beta$1/2 exposure.
Conversely, elevated E-cadherin expression depended upon the activity of NFAT transcription factors downstream of VEGF-A stimulation.
To further isolate the role of NFAT on this hybrid state, we simulated the inhibition of NFAT transcriptional activity across all conditions (all else being equal).
NFAT inhibition in combination with VEGF-A and TGF$-\beta$1/2 treatments blocks increased activation of E-cadherin (Fig. \ref{fg:F3}D and Fig. \ref{fg:F3}E).
Lastly, NFATc inhibition in combination with simultaneous TGF$-\beta$1/2 and VEGF-A exposure repressed nearly all E-cadherin expression, shifting nearly the entire population towards a mesenchymal phenotype (Fig. \ref{fg:F4}F). Taken together, high levels of nuclear localized phosphorylated Sp1 correlated with Vimentin expression, while NFATc transcriptional activity was predicted to be critical for maintaining E-cadherin expression in the prescence of competing signals.
In particular, when present, VEGF-A signaling through NFATc was predicted to provide a protective up-regulation of E-cadherin.

@@

\section*{Materials and Methods}
All model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.

@@

Our signaling network reconstruction was based on Medici et al. who identified the pathways though which MDCKII, DLD1 colon carcinoma,
and A375 melanoma cells transition towards a mesenchymal phenotype \citep{Medici:2008fk}.
Sequential activation of MAPK and Smad pathways were initiated upon addition of TGF$-\beta$1/2.  Briefly, TGF$-\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \citep{Medici:2006qa}.  Snail, a known repressor of junctional proteins, inhibits the expression of E-cadherin \citep{Cano:2000kh}. This initial repression of E-cadherin leads to a release of $\beta$-catenin from the cell membrane. This release of $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$-\beta$3 expression \citep{Medici:2008fk}. The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \c\
itep{Medici:2008fk}. GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway
\citep{Zhou:2004kx,Larue:2005vn}. Thereby, further $\beta$-catenin release also resulted from by TGF$-\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \citep{Derynck:2003fc}. This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \citep{Massague:2005qc}. In the model, receptors are represented as either bound or unbound complexes.
Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers and translocate to the nucleus. pSmads complexes up-regulate other transcription factors, such as LEF-1. The pSmad2/4-LEF-1 complex has been shown to directly repress the E-cadherin gene \citep{Nawshad:2007kl}. LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \citep{Medici2011}.
In the model, Smad signaling is lumped into a single smad species that can act in a co-dependent fashion with LEF1 to downregulate E-cadherin.
The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors.
Binding of extracellular TGF$-\beta$1/2 with TGF$-\beta$ surface receptors I/II (TGF$-\beta$R-I/II) initiates the assembly of adapter complexes which starts the
downstream signaling program. In the model, TGF$-\beta$1/2 binds TGF$-\beta$R-I/II followed by the recruitment of activin receptor-like kinase 1 (ALK1) and
TGF$-\beta$ surface receptor III (TGF$-\beta$R-III) to form the activated receptor complex \citep{Derynck:2003fc}.
Alternatively, we also included activin receptor-like kinase 5 (ALK5) recruitment in combination with Endoglin and TGF$-\beta$R-III
as a second (redundant) activated receptor complex \citep{Gatza:2010if}.
Complex assembly activates the serine/threonine kinase activity on the receptor, leading to the recruitment and phosphorylation of Smad partners \citep{Massague:2005qc}.
Phosphorylated Smads2/3 (pSmad2/3) form heterodimers with partner Smad4 and then translocate to the nucleus where they act as both transcriptional activators and repressors.
Nuclear pSmad2/3-Smad4 form transcriptional complexes with several genes in the model including lymphoid enhancer-binding factor 1 (\emph{LEF-1}), Nuclear factor of activated T-cells, cytoplasmic 1 (\emph{NFACT1}), and Specificity Protein 1 (\emph{SP1}).  On the other hand, nuclear pSmad2/3-Smad4 represses (in combination with the LEF-1 protein) the expression of E-cadherin (\emph{Cdh1}) \citep{Nawshad:2007kl} and Cadherin 5, type 2 (VE-Cadherin encoded by \emph{Cdh5}).
Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \citep{Cano:2000kh}.
However, this transition is not solely driven by transcriptional events.
At the protein level, the repression of E-cadherin leads to a release of $\beta$-catenin from cell membrane.
Cystolic $\beta$-catenin then translocates to the nucleus and forms transcriptionally-active complexes with immunoglobulin transcription factor 2 (TCF-4)
to drive TGF$-\beta$3 expression \citep{Medici:2008fk}. The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}. GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway
\citep{Zhou:2004kx,Larue:2005vn}. Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \citep{Suehiro:2014aa},
as supported by our VEGF-A experimental data (Fig. \ref{fg:VEGFA-Data}).

>>>>

The predictive power of the ensemble was tested using both cross validation and by comparing simulations with data sets not used for model training.
In whole, 100\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
generated from a random starting point. Conversely, we \emph{predicted} 100\% of the training objectives, at a 95\% confidence interval compared to randomized parameters (Wicoxon non-parametric test).
The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 (not used for model training) to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L). Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
A listing of objective function values resulting from training, cross validation and the random parameter control is given in the supplement (Fig. \ref{fg:ObjTable}).

@@

\subsection*{TGF$-\beta$1/2 and VEGF-A exposure promotes phenotype heterogeneity through NFATc and phosphorylated Sp1}
While we captured the central tendency of many of the molecular features of EMT induction following TGF$-\beta$1/2 exposure, an often neglected but important emergent
feature of developmental and pathological programs is population heterogeneity \citep{Park:2010fu}.
We (and others) have previously hypothesized that deterministic model ensembles can simulate population behavior, at least at a course grained level \citep{Lequieu:2011fj}.
We tested this hypothesis by analyzing the response of the population of EMT models to extracellular cues and then comparing this response to flow cytometry studies.
We quantified the phenotypic response of the individual members of the ensemble to TGF$-\beta$1/2 stimulation for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial) following the addition of TGF$-\beta$1/2 alone (Fig. \ref{fg:F3}), and/or VEGF-A in combination with NFATc inhibitors (Fig. \ref{fg:F3}).
%We were further able to use this analysis to examine the activity and effects of underlying signaling molecules.

We identified model subpopulations that exhibited different behaviors following exposure to TGF$-\beta$1/2 (Fig. \ref{fg:F3}B).
Analysis of the molecular signatures of these subpopulations suggested the abundance, localization and state of the Sp1, AP-1 and NFATc transcription factors controlled population heterogeneity.
The majority of models ($>$80\%) responded to treatment, moving away from the untreated population (Fig \ref{fg:F3}A-C, gray).
These models showed the classically expected behavior, a switch from an epithelial to mesenchymal phenotype following TGF$-\beta$1/2 exposure.
Some models resembled untreated cells; they had evaluated nuclear localized phosphorylated Sp1, relative to non-induced cells,
which decreased E-cadherin expression through Slug-mediated inhibition, which in turn increased Vimentin expression through TGF$-\beta$3 autocrine signaling and the liberation of $\beta$-catenin. However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (\ref{fg:F3}C, black arrow), but also present in a small percentage of untreated cells (Fig. \ref{fg:F3}A).
Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in simultaneously increased Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).

%Analysis of these hypothetical cells suggested they had deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression.

To better understand the hybrid phenotype, we simulated the response of the model population to TGF$-\beta$1/2 and VEGF-A treatment with and without NFATc inhibitors (Fig. \ref{fg:F3}).
As expected, stimulation with VEGF-A (50 a.u.) maintained an epithelial population (Fig. \ref{fg:F3}A), while TGF$-\beta$1/2 (10 a.u.) exposure shifted
the population from an epithelial to a mesenchymal phenotype (Fig. \ref{fg:F3}B).
On the other hand, combined stimulation with TGF$-\beta$1/2 (10 a.u.) and VEGF-A (50 a.u.) increased both E-cadherin and Vimentin expression, resulting in a hybrid phenotype with both epithelial and mesenchymal characteristics (Fig. \ref{fg:F3}C).
Vimentin expression was correlated with high levels of nuclear phosphorylated Sp1, following TGF$-\beta$1/2 exposure.
Conversely, elevated E-cadherin expression depended upon the activity of NFAT transcription factors downstream of VEGF-A stimulation.
To further isolate the role of NFAT on this hybrid state, we simulated the inhibition of NFAT transcriptional activity across all conditions (all else being equal).
NFAT inhibition in combination with VEGF-A or TGF$-\beta$1/2 treatments blocked increased E-cadherin expression in the case of VEGF-A (Fig. \ref{fg:F3}D), but did not influence TGF$-\beta$1/2 signaling (Fig. \ref{fg:F3}E).
Lastly, NFATc inhibition in combination with simultaneous TGF$-\beta$1/2 and VEGF-A exposure repressed nearly all E-cadherin expression, shifting nearly the entire population towards a mesenchymal phenotype (Fig. \ref{fg:F3}F). Taken together, high levels of nuclear localized phosphorylated Sp1 correlated with Vimentin expression,
while NFATc transcriptional activity was critical for maintaining E-cadherin expression in the presence of competing signals.

%In particular, when present, VEGF-A signaling through NFATc was predicted to provide a protective up-regulation of E-cadherin.

@@

\section*{Materials and Methods}
The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.

@@

Our signaling network reconstruction was based on Medici et al. who identified the pathways though which MDCKII, DLD1 colon carcinoma,
and A375 melanoma cells transition towards a mesenchymal phenotype \citep{Medici:2008fk}.
Sequential activation of MAPK and Smad pathways were initiated upon addition of TGF$-\beta$1/2.  Briefly, TGF$-\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \citep{Medici:2006qa}.  Snail, a known repressor of junctional proteins, inhibits the expression of E-cadherin \citep{Cano:2000kh}. This initial repression of E-cadherin leads to a release of $\beta$-catenin from the cell membrane. This release of $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$-\beta$3 expression \citep{Medici:2008fk}. The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \c\
itep{Medici:2008fk}. GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway
\citep{Zhou:2004kx,Larue:2005vn}. Thereby, further $\beta$-catenin release also resulted from by TGF$-\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \citep{Derynck:2003fc}. This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \citep{Massague:2005qc}. In the model, receptors are represented as either bound or unbound complexes.
Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers and translocate to the nucleus. pSmads complexes up-regulate other transcription factors, such as LEF-1. The pSmad2/4-LEF-1 complex has been shown to directly repress the E-cadherin gene \citep{Nawshad:2007kl}. LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \citep{Medici2011}.
In the model, Smad signaling is lumped into a single smad species that can act in a co-dependent fashion with LEF1 to downregulate E-cadherin.
The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors.
Binding of extracellular TGF$-\beta$1/2 with TGF$-\beta$ surface receptors I/II (TGF$-\beta$R-I/II) initiates the assembly of adapter complexes which starts the
downstream signaling program.
Complex assembly activates the serine/threonine kinase activity on the receptor, leading to the recruitment and phosphorylation of Smad partners \citep{Massague:2005qc}.
Phosphorylated Smads2/3 (pSmad2/3) form heterodimers with partner Smad4 and then translocate to the nucleus where they act as both transcriptional activators and repressors.
Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \citep{Cano:2000kh}.
However, this transition is not solely driven by transcriptional events.
At the protein level, the repression of E-cadherin leads to a release of $\beta$-catenin from cell membrane.
Cystolic $\beta$-catenin then translocates to the nucleus and forms transcriptionally-active complexes with immunoglobulin transcription factor 2 (TCF-4)
to drive TGF$-\beta$3 expression \citep{Medici:2008fk}. The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}. GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway.
\citep{Zhou:2004kx,Larue:2005vn}. Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \citep{Suehiro:2014aa},
as supported by our VEGF-A experimental data (Fig. \ref{fg:VEGFA-Data}).
